Dancing on Water: The Choreography of Sulfur Dioxide Adsorption to Aqueous Surfaces

### 2011-07-04 Conclusions.tex
fixed x-axis location on figures. Redid conclusions. Minor edits for the rest.

Presented herein are the results of several classical molecular dynamic simulations that focus on understanding how surface water molecules and adsorbing \suldiox~gas molecules twist and turn as the gas transits the interfacial region. The computational studies emulate and expand on the experimental spectroscopic studies from this laboratory which have found \suldiox~surface complexation at a water surface.\cite{Tarbuck2005,Tarbuck2006,Ota2011}  These spectroscopic studies show clear evidence of \suldiox-water surface complexation but details about this surface complex could only be inferred from spectral changes in the surface water spectrum since \suldiox~could not be monitored directly.   These simulations do not have that limitation and hence can provide information about the behavior of both surface partners and in particular, how their proximity influences the orientation behavior of each other.  The orientational information obtained in these simulations are provided via calculated depth profiles which show the molecular distribution of orientations of the two different interfacial molecules throughout the dimensions of the interfacial region.

@@

Through these simulations we characterize the orientational behavior of \suldiox~during and after adsorption. Our equilibrium neat-water simulations show that a single \suldiox~molecule, representing a low concentration, has a high surface affinity. At a high \suldiox~concentration in the saturated systems, \suldiox~molecules are also surface active, and are found further out of the water phase than at the lower concentration. These \suldiox~form a bound layer of molecules that crowd the surface and interact with the surface waters. The orientation of \suldiox~on the water surface was found to be similar for both low and high concentrations. Those \suldiox~molecules at or below the surface strongly orient with the sulfur atom pointed in towards the water bulk, and the oxygen atoms out towards the gas phase. The \suldiox~slightly above the water surface loses the orientational preference within 10 $\AA$ and those further from the water are more isotropically oriented. Figure \ref{fig:so2-surface-cartoon} depicts what the neat-water and saturated surfaces might look like for both \suldiox~and \wat~orientations and locations.

@@

Steered molecular dynamics simulations were used to model the behavior of an adsorbing \suldiox~as it moves from the gas phase above the water down through the surface and into the bulk. The results for the transit through the interface show that in both systems of low and high \suldiox~concentration an adsorbing \suldiox~has very similar orientation to those already bound to the water surface. The \suldiox~reorients as it makes its first contact with the water interface. Within 5 $\AA$ of the surface the \suldiox~is mostly oriented with its sulfur towards the water phase. The \suldiox~pulled further into the water bulk retains its orientation until it is past the interfacial region and then isotropically orients with the bulk water.

Having now examined the transition of \suldiox~from the gas to adsorbed aqueous phases, we are better able to research the many questions remaining of adsorption and interfacial chemistry of gas molecules. What species form during the adsorption transition, and how do molecular properties affect the process? Can we form a theory that will more fully explain the transit of many small molecules of environmental and industrial interest? This study is one of several characterizing \suldiox~adsorption and behavior on aqueous surfaces. We plan to report further results of ongoing computational simulation and experimental studies regarding temperature and chemical constituent effects on atmospherically relevant water surfaces.

>>>>

Presented herein are the results of several classical molecular dynamic simulations that focus on understanding how surface water molecules and adsorbing \suldiox~gas molecules twist and turn as the gas adsorbs and transits the interfacial region. The computational studies emulate and expand on the experimental spectroscopic studies from this laboratory which have found \suldiox~surface complexation at a water surface.\cite{Tarbuck2005,Tarbuck2006,Ota2011}  These spectroscopic studies show clear evidence of \suldiox-water surface complexation but details about this surface complex could only be inferred from spectral changes in the surface water spectrum since \suldiox~could not be monitored directly.   These simulations do not have that limitation and hence can provide information about the behavior of both surface partners and in particular, how their proximity influences the orientation behavior of each other.  The orientational information obtained in these simulations are provided via calculated depth profiles which show the molecular distribution of orientations of the two different interfacial molecules throughout the dimensions of the interfacial region.

@@

Through these simulations we also characterize the orientational behavior of \suldiox~during and after adsorption. Our equilibrium neat-water simulations show that a single \suldiox~molecule, representing a low concentration, has a high surface affinity. At a high \suldiox~concentration in the saturated systems, \suldiox~molecules are also surface active, and are found further out of the water phase than at the lower concentration. These \suldiox~molecules form a bound layer that crowds the surface and interacts with the surface waters. The orientation of \suldiox~on the water surface was found to be similar for both low and high concentrations. Those \suldiox~molecules at or below the surface water location strongly orient with the sulfur atom pointed in towards the water bulk, and the oxygen atoms out towards the gas phase. The \suldiox~molecules slightly above the water surface lose this net orientation within 10 \angs. Those  molecules further from the water are more isotropically oriented. Figure \ref{fig:so2-surface-cartoon} depicts what the neat-water and saturated surface molecules look like for both \suldiox~and \wat~orientations and locations based on the calculations.

@@

Steered molecular dynamics simulations were used to model the behavior of an adsorbing \suldiox~as it moves from the gas phase above the water down through the surface and into the bulk. The \suldiox~reorients as it makes its first contact with the water interface. Within 5 \angs~of the surface the \suldiox~is mostly oriented with its sulfur towards the water phase. The results for the transit through the interface show that in both systems of low and high \suldiox~concentration an adsorbing \suldiox~near the interface has very similar orientation to those molecules already bound to the water surface. The \suldiox~pulled further into the water bulk retains its orientation until it is past the interfacial region and then isotropically orients with the bulk water.

These studies provide a starting point for future studies in this area that seek to understand how gases of different concentrations and chemical composition adsorb and transit across an aqueous/air interface. Obtaining such knowledge will be invaluable for understanding many environmental aerosol and land water systems where gaseous uptake at a water surface does not conform to expectations. In such cases, surface complexes are often invoked with little direct evidence of the formation of such complexes. Investigations of these and related low-temperature effects will be forthcoming in future publications.

%Having now examined the transition of \suldiox~from the gas to adsorbed aqueous phases, we are better able to research the many questions remaining of adsorption and interfacial chemistry of gas molecules. What species form during the adsorption transition, and how do molecular properties affect the process? Can we form a theory that will more fully explain the transit of many small molecules of environmental and industrial interest? This study is one of several characterizing \suldiox~adsorption and behavior on aqueous surfaces. We plan to report further results of ongoing computational simulation and experimental studies regarding temperature and chemical constituent effects on atmospherically relevant water surfaces.

### 2011-07-04 Introduction.tex
fixed x-axis location on figures. Redid conclusions. Minor edits for the rest.

An important gas for developing a picture of gaseous adsorption and entry into a water surface is \suldiox.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011} Sulfur dioxide enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well.

In this work we provide a molecular picture of \suldiox~adsorption on a water surface.  This study demonstrates the strong orientational effect of surface water molecules on adsorbing gas during the approach and entry into the surface region at both high and low \suldiox~surface concentrations.  Obtaining a more complete picture of this adsorption process is important for understanding gaseous adsorption of this environmentally important gas on water and aerosol surfaces as well as being a model system for understanding the more general nature of gases at aqueous interfaces.

These computational studies complement and significantly expand the picture developed in our recent experimental vibrational sum frequency spectroscopy (VSFS) studies of \suldiox~adsorption of aqueous solutions of various compositions and temperatures,\cite{Tarbuck2005,Tarbuck2006} and the subsequent studies using both classical and ab initio simulations.  These experimental studies showed that an \suldiox~surface hydrate complex forms when an aqueous surface is exposed to \suldiox~gas. The computational study by Baer et al.\cite{Baer2010} then made a series of predictions of the specific nature of the hydrated complex through classical and ab initio simulations. That work developed a detailed picture of the nature of the \suldiox~surface complex with water, and related it to the surface water OH vibrational IR spectra.  The most recent experimental studies have shown that whereas the binding of gaseous \suldiox~to a water surface is greatly enhanced at cold temperatures, the reversibility of the adsorption process remains.  Complimentary experiments showed that low pH aqueous environments inhibit the bulk reactions of \suldiox, but do not affect the surface binding or its reversibility.  What is apparent in the VSF spectra obtained in all of these experiments is the tendency of water to reorient upon surface bonding, with the effect becoming more pronounced at high \suldiox~surface concentrations. Since the \suldiox~molecule was not specifically probed, conclusions on how \suldiox~bonding contributes to reorienting surface water molecules and the orientation of \suldiox~itself upon approach and surface bonding could only be inferred.

>>>>

An important gas for developing a picture of gaseous adsorption and entry into a water surface is sulfur dioxide.\cite{Lattanzi2010,Shah2011,Tzivian2011,Johns2011,Faloona2009,Jurkat2010,Wu2011} \suldiox~enters the environment as an important industrial product, and also naturally through terrestrial processes. Atmospheric dust particles and gases have been implicated in the oxidation of \suldiox, and act as reaction surfaces for chemical mechanisms that are still poorly understood.\cite{Baltrusaitis2011,Rubasinghege2010,Li2007} \suldiox~acts as a major component of atmospheric pollution, and is a precursor to acid rain formation, and cloud nucleation. Its high solubility in water makes \suldiox~an integral compound in many aqueous atmospheric reactions, as well.  Obtaining a more complete picture of the \suldiox~adsorption process is important for understanding gaseous adsorption of this environmentally important gas on water and aerosol surfaces as well as being a model system for understanding the more general nature of gases at aqueous interfaces.

In this work we provide a molecular picture of \suldiox~adsorption on a water surface.  This study demonstrates the strong orientational effect of surface water molecules on the adsorbing gas during the approach and entry into the surface region at both high and low \suldiox~surface concentrations.  These computational studies complement and significantly expand the picture developed in our recent experimental vibrational sum frequency spectroscopy (VSFS) studies of \suldiox~adsorption of aqueous solutions of various compositions and temperatures,\cite{Tarbuck2005,Tarbuck2006} and the subsequent studies using both classical and ab initio simulations.  These experimental studies showed that an \suldiox~surface hydrate complex forms when an aqueous surface is exposed to \suldiox~gas. The computational study by Baer et al.\cite{Baer2010} then made a series of predictions of the specific nature of the hydrated complex through classical and ab initio simulations. That work developed a detailed picture of the nature of the \suldiox~surface complex with water, and related it to the surface water OH vibrational IR spectra.  The most recent experimental studies have shown that whereas the binding of gaseous \suldiox~to a water surface is greatly enhanced at cold temperatures, the reversibility of the adsorption process remains.\cite{Ota2011}  Complementary experiments showed that low pH aqueous environments inhibit the bulk reactions of \suldiox, but do not affect the surface binding or its reversibility.  What is apparent in the VSF spectra obtained in all of these experiments is the tendency of water to reorient upon surface bonding, with the effect becoming more pronounced at high \suldiox~surface concentrations. Since the \suldiox~molecule was not specifically probed, conclusions on how \suldiox~bonding contributes to reorienting surface water molecules and the orientation of \suldiox~itself upon approach and surface bonding could only be inferred.

### 2011-07-04 Methods.tex
fixed x-axis location on figures. Redid conclusions. Minor edits for the rest.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.

This method for finding the outer monolayer location was implemented, and the surface location is plotted as a function of simulation time in Figure \ref{fig:density-flaw}B. It is apparent from the surface location plot that the water slab location, and thus the surface location, drifts over the 10 ns, spanning approx. 12 $\AA$. However, the maximum standard deviation of the positions of the waters comprising the surface layer at each time step is only 1.85 $\AA$. Consequently, all depth locations in our analyses are calculated relative to the instantaneous surface location at the corresponding time step of the trajectories.

@@

%\subsection{Hydrate Complex Analysis}

%During the adsorption process of a gaseous \suldiox~moving in towards a water surface hydrate complexes are formed as the waters bind to the molecule. Computationally, there are various methods to determine hydrate complex formation. The most simple metric is to count the number of hydrogen bonds formed from the \suldiox~to neighboring molecules. This method can be furthered extended to also include the strength of those bonding interactions by additionally reporting the bondlengths to the hydrogen bonding partners. Longer bondlengths are typical of weaker bonds, and this method is used in this work to identify the bonding environment of the \suldiox~during solvation.

%A simple count of hydrogen bonds is telling of the interactions and environments of an adsorbing \suldiox. However, a more sophisticated analysis is needed to fully characterize the geometry of any complexes that form. In this work a graph theoretical method has been used to identify hydrate complexes that form cyclic structures in addition to those that form more simple hydrogen bonding geometries. Previously published works have detailed graph theory for use in molecular systems, and have used graph theoretical methods to describe ice and water clusters, structural energies, as well as in the study of relationships in biological systems.\cite{Huber2007,Shi2005,Radhakrishnan1991,Anick2002} As such, we will not present here a review of the basic graph theoretical details that are well documented in the referenced works. As shown in Figure \ref{fig:so2-complex-graph}, the atoms of a \suldiox~and its nearest neighbors are represented as a set of verrtices, with the covalent and hydrogen bonds between them forming the edges of a graph.

%We present here a technique for determining bonding cycles in a hydrate complex, but this technique is general enough to be applied to any graph representation of simple molecular systems. The first step in the determination of cycles is translating the atoms (vertices) and bonds (edges) to a graph. Figure \ref{fig:so2-complex-graph} shows the procedure visually, where here the root vertex of the graph is the \suldiox-sulfur, and each atom is numbered by the distance, in atoms, to the root vertex. We use only the closest neighboring water molecules that may take part in a cyclic complex in building the graph. After forming the graph, a breadth-first search (BFS) is performed starting at the root vertex.\cite{Knuth1997} The BFS algorithm uses a coloring technique where all unexamined vertices are colored white, queued vertices are gray, and examined vertices are black. During the BFS, encountering a gray vertex as a target of an adjacent edge during examination indicates that a cycle has been found in the graph. At each vertex dequeue for examination, the parent of every vertex is recorded such that the lineage from gray target to root vertex can be reconstructed to determine the participating atoms in the cycle.

%\begin{figure}[h!]
%\begin{center}
%\includegraphics[scale=1.0]{images/Cycles/so2complexfigure.png}
%\caption{A \suldiox~and its nearest neighbors are represented as a graph of vertices and edges. The bonds are all given equal weights and the graph is undirected. The numbering shows the relative distance from any atom to the root node (\suldiox-sulfur 1), and also enumerates the iteration through the BFS process at which the atom is discovered. A BFS on this graph would result in discovering the doubly-bound hydrogen (5) as a gray target, and the gray source would be either of the connected oxygens (4). At each iteration the predecessor in the BFS is recorded in order to reconstruct the connectivity of the cycle. This cycle is composed of 8 atoms, and 3 molecules}
%\label{fig:so2-complex-graph}
%\end{center}
%\end{figure}

>>>>

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.

This method for finding the outer monolayer location was implemented, and the surface location is plotted as a function of simulation time in Figure \ref{fig:density-flaw}B. It is apparent from the surface location plot that the water slab location, and thus the surface location, drifts over the 10 ns, spanning approximately 12 \angs. However, the maximum standard deviation of the positions of the waters comprising the surface layer at each time step is only 1.85 \angs. Consequently, all depth locations in our analyses are calculated relative to the instantaneous surface location at the corresponding time step of the trajectories.

@@
